---
title: "Noncentrality of IMPC summary statistics"
author: "Donghyung Lee"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{IMPC summary stat exploratory data analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE)
```

## Load R packages
```{r load_packages, message=FALSE}
rm(list=ls())
library(plyr)
library(ggplot2)
library(RColorBrewer)
library(reshape2)
library(mvtnorm)
library(corrplot)
library(data.table) #fread
library(plotly)
library(pheatmap)
library(heatmaply)
library(kableExtra)

color.vec <- brewer.pal(8, "Set1")


FIQT <- function(z=z, min.p=10^-300){
    pvals<-2*pnorm(abs(z),low=F)
    pvals[pvals<min.p]<- min.p
    adj.pvals<-p.adjust(pvals,method="fdr")
    mu.z<-sign(z)*qnorm(adj.pvals/2,low=F)
    mu.z[abs(z)>qnorm(min.p/2,low=F)]<-z[abs(z)>qnorm(min.p/2,low=F)]
    mu.z
}

```

## Reading the summary data
```{r load_datasets, echo=TRUE}
## load IMPC summary stats 7.0
IMPC.summary.stats.file <- "~/Dropbox/KOMP/results/IMPC_Summary_Stat_Study/R_input/version_7.0/IMPC_ALL_statistical_results.csv"
IMPC.summary.stats <- as.data.frame(fread(IMPC.summary.stats.file, header=TRUE, sep=","))
dim(IMPC.summary.stats)
```

## Delete phenotypes with less than 100 summary stats
```{r prep_KOMP}

IMPC.summary.stats$marker_symbol_combined <- paste0(IMPC.summary.stats$marker_symbol,"_",IMPC.summary.stats$zygosity)

summ.data <- IMPC.summary.stats

summ.data <- summ.data[,c("procedure_name","parameter_name","marker_symbol","marker_symbol_combined",
                        "genotype_effect_parameter_estimate","genotype_effect_stderr_estimate",
                        "genotype_effect_p_value")]
summ.data$zscore <- (summ.data$genotype_effect_parameter_estimate/summ.data$genotype_effect_stderr_estimate) # Z-score
summ.data$phenotype <- paste0(summ.data$procedure_name,"_",summ.data$parameter_name)
summary(summ.data$zscore)
dim(summ.data)
summ.data <- na.omit(summ.data)
summ.data$fdr <- p.adjust(summ.data$genotype_effect_p_value, method="BH",
                                                                       n= length(na.omit(summ.data$genotype_effect_p_value)))
dim(summ.data)

# filter out phenotypes
pheno.names <- as.data.frame.vector(table(summ.data$phenotype))
colnames(pheno.names) <- "Counts"
pheno.list <- rownames(pheno.names)[pheno.names$Counts>=100]
summ.data <- summ.data[summ.data$phenotype%in%pheno.list,]
dim(summ.data)

sdf <- ddply(summ.data, c("procedure_name","phenotype")
             , summarise,
             mean=mean(zscore, na.rm=TRUE), sd=sd(zscore, na.rm=TRUE), 
             n= sum(!is.na(zscore)), 
             #tstat=t.test(zscore)$statistic, 
             #pval=t.test(zscore)$p.value, 
             num.sig.genes = sum(fdr<0.05, na.rm=TRUE))

sdf <- sdf[order(sdf$mean),]
summ.data$phenotype <- factor(summ.data$phenotype, levels=sdf$phenotype)
kable(sdf) %>%
  kable_styling(bootstrap_options = c("striped","condensed","responsive"), full_width = F, font_size = 12) %>%
  scroll_box(height = "500px")
```

## Z-score distribution of each phenotype
```{r plot1,  fig.height=50, fig.width=10}
p <- ggplot(summ.data, aes(phenotype, zscore, title=""))
p <- p + geom_boxplot()
p <- p + geom_hline(yintercept = 0, color="red")
p <- p + geom_hline(yintercept = -2, color="blue") ## z=1.96 for two-sided test at alpha=0.05
p <- p + geom_hline(yintercept = 2, color="blue")  ## z=-1.96 for two-sided test
p <- p + ggtitle("Association Z-score Non-centrality")
p <- p + xlab("Domain & Phenotype")
p <- p + ylab("Phenotype-Genotype Association Z-score")
p <- p + coord_flip()
p
```

## The number of significant mutant lines per phenotype as a function of Z-score mean
```{r plot2, fig.height=8, fig.width=12}

p <- ggplot(sdf, aes(x=mean, y=num.sig.genes, text = phenotype)) 
p <- p + geom_point(size=2, aes(colour = procedure_name))
p <- p + geom_smooth(method = "loess", size=1, aes(group=1))
p <- p + labs(y= "# of Sig Genes", x= "Z-score Mean Per Phenotype")
p <- p + theme_bw()
p <- p + theme(plot.background = element_blank() ,
                              panel.border = element_blank(),
                              panel.background = element_blank())
p <- p + theme(axis.line = element_line(color = 'black'))
p <- p + theme(axis.title.x = element_text(size = 15, vjust= 0))
p <- p + theme(axis.title.y = element_text(size = 15, vjust= 0))
p <- p + theme(strip.text.x = element_text(size = 8))
ggplotly(p)

```

## Proportion of sig. mutant lines per phenotype as a function of Z-score mean per phenotype
```{r plot3, fig.height=8, fig.width=12}

p <- ggplot(sdf, aes(x=mean, y=100*sdf$num.sig.genes/sdf$n, text = phenotype)) 
p <- p + geom_point(size=2, aes(colour = procedure_name))
p <- p + geom_smooth(method = "loess", size=1, aes(group=1))
p <- p + labs(y= "Prop. Sig Genes", x= "Z-score Mean Per Phenotype")
p <- p + theme_bw()
p <- p + theme(plot.background = element_blank() ,
                              panel.border = element_blank(),
                              panel.background = element_blank())
p <- p + theme(axis.line = element_line(color = 'black'))
p <- p + theme(axis.title.x = element_text(size = 15, vjust= 0))
p <- p + theme(axis.title.y = element_text(size = 15, vjust= 0))
p <- p + theme(strip.text.x = element_text(size = 8))

ggplotly(p)

```

## Phenotypic correlation estimated using association Z-scores
```{r plot_correlation,  fig.height=40, fig.width=40}

summ.zscore.df <- summ.data[,c("procedure_name","phenotype","marker_symbol_combined","zscore")]
dim(summ.zscore.df)
summ.zscore.df <- dcast(data=summ.zscore.df, formula= phenotype~marker_symbol_combined, fun.aggregate = sum, value.var = "zscore")
dim(summ.zscore.df)

summ.zscore.mat <- as.matrix(summ.zscore.df[,-1])
rownames(summ.zscore.mat) <- summ.zscore.df$phenotype
dim(summ.zscore.mat)

pheno.cor.mat <- cor(t(summ.zscore.mat), use="pairwise")
pheatmap(pheno.cor.mat)
```


## Delete mutantlines with less than 100 summary stats
```{r prep_KOMP2}

summ.data <- IMPC.summary.stats

summ.data <- summ.data[,c("procedure_name","parameter_name","marker_symbol","marker_symbol_combined",
                        "genotype_effect_parameter_estimate","genotype_effect_stderr_estimate",
                        "genotype_effect_p_value")]
summ.data$zscore <- (summ.data$genotype_effect_parameter_estimate/summ.data$genotype_effect_stderr_estimate) # Z-score
summ.data$phenotype <- paste0(summ.data$procedure_name,"_",summ.data$parameter_name)
summary(summ.data$zscore)
dim(summ.data)
summ.data <- na.omit(summ.data)
summ.data$fdr <- p.adjust(summ.data$genotype_effect_p_value, method="BH",
                                                                       n= length(na.omit(summ.data$genotype_effect_p_value)))
dim(summ.data)


# filter out mutant lines
mutantlines <- as.data.frame.vector(table(summ.data$marker_symbol_combined))
colnames(mutantlines) <- "Counts"
mutantline.list <- rownames(mutantlines)[mutantlines$Counts>100]
length(mutantline.list)
summ.data <- summ.data[summ.data$marker_symbol_combined%in%mutantline.list,]
dim(summ.data)

sdf <- ddply(summ.data, c("marker_symbol","marker_symbol_combined")
             , summarise,
             mean=mean(zscore, na.rm=TRUE), sd=sd(zscore, na.rm=TRUE), 
             n= sum(!is.na(zscore)), 
             num.sig.phenotypes = sum(fdr<0.05, na.rm=TRUE))

sdf <- sdf[order(sdf$mean),]
summ.data$marker_symbol_combined <- factor(summ.data$marker_symbol_combined, levels=sdf$marker_symbol_combined)
kable(sdf) %>%
  kable_styling(bootstrap_options = c("striped","condensed","responsive"), full_width = F, font_size = 12) %>%
  scroll_box(height = "500px")
```

## Z-score distribution of each mutant line
```{r plot_mean,  fig.height=50, fig.width=10}
p <- ggplot(summ.data, aes(marker_symbol_combined, zscore, title=""))
p <- p + geom_boxplot()
p <- p + geom_hline(yintercept = 0, color="red")
p <- p + geom_hline(yintercept = -2, color="blue") ## z=1.96 for two-sided test at alpha=0.05
p <- p + geom_hline(yintercept = 2, color="blue")  ## z=-1.96 for two-sided test
p <- p + ggtitle("Association Z-score Non-centrality")
p <- p + xlab("Knockout Gene")
p <- p + ylab("Phenotype-Genotype Association Z-score")
p <- p + coord_flip()
p
```

## The number of significant phenotypes as a function of Z-score mean of each mutant line
```{r plot5,  fig.height=6, fig.width=8}

p <- ggplot(sdf, aes(x=mean, y=100*num.sig.phenotypes/n, text = marker_symbol_combined)) 
p <- p + geom_point(size=2)
p <- p + geom_smooth(method = "loess", size=1, aes(group=1))
p <- p + labs(y= "Prop. Sig Phenotypes", x= "Z-score Mean Per Gene")
p <- p + theme_bw()
p <- p + theme(plot.background = element_blank() ,
                              panel.border = element_blank(),
                              panel.background = element_blank())
p <- p + theme(axis.line = element_line(color = 'black'))
p <- p + theme(axis.title.x = element_text(size = 15, vjust= 0))
p <- p + theme(axis.title.y = element_text(size = 15, vjust= 0))
p <- p + theme(strip.text.x = element_text(size = 8))
ggplotly(p)

```

